Aligning species range data to better serve science and conservation

### 2015-12-29 ms_5_refs.Rmd

#References

@@

Kaschner, K., D.P. Tittensor, J. Ready, T Gerrodette and B. Worm (2011). Current and Future Patterns of Global Marine Mammal Biodiversity. PLoS ONE 6(5): e19653. PDF

Ready, J., K. Kaschner, A.B. South, P.D Eastwood, T. Rees, J. Rius, E. Agbayani, S. Kullander and R. Froese (2010). Predicting the distributions of marine organisms at the global scale. Ecological Modelling 221(3): 467-478. PDF

__Papers based on AquaMaps:__ ___Use these for section 4?___

Jones, M.C., S.R. Dyeb, J.K. Pinnegar and W.W.L. Cheung (2012). Modelling commercial fish distributions: Prediction and assessment using different approaches. Ecological Modelling 225(2012): 133-145. PDF

Coll, M., C. Piroddi, J. Steenbeek, K. Kaschner, F. Ben Rais Lasram et al. (2010). The biodiversity of the Mediterranean Sea: estimates, patterns, and threats. PLoS ONE 5(8): e11842. PDF

Martin C.S., Fletcher R., Jones M.C., Kaschner K., Sullivan E., Tittensor D.P., Mcowen C., Geffert J.L., van Bochove J.W., Thomas H., Blyth S., Ravillious C., Tolley M., Stanwell-Smith D. (2014). Manual of marine nd coastal datasets of biodiversity importance. May 2014 release. Cambridge (UK): UNEP World Conservation Monitoring Centre. 28 pp. (+ 4 annexes totalling 174 pp. and one e-supplement). PDF

@@

Hurlbert 2007

Jetz 2008

Pimm 2014

Rondinini 2006

@@

<!--
Halpin, P.N., A.J. Read, E. Fujioka, B.D. Best, B. Donnelly, L.J. Hazen, C. Kot, K. Urian, E. LaBrecque, A. Dimatteo, J. Cleary, C. Good, L.B. Crowder, and K.D. Hyrenbach. 2009. OBIS-SEAMAP: The world data center for marine mammal, sea bird, and sea turtle distributions. Oceanography 22(2):104-115
-->

@@

* _IUCN reference:_ 
    * https://www.conservationtraining.org/course/view.php?id=217&lang=en
    * http://www.iucnredlist.org/technical-documents/red-list-training/iucnspatialresources

>>>>

# References

@@

Kaschner, K., R. Watson, A. W. Trites, D. Pauly (2006). Mapping world-wide distributions of marine mammal species using a relative environmental suitability (RES) model.  Marine Ecology Progress Series 316: 285–310. _check this citation journal name..._ _This outlines the basic RES methodology - AM development_

Kaschner, K., D.P. Tittensor, J. Ready, T Gerrodette and B. Worm (2011). Current and Future Patterns of Global Marine Mammal Biodiversity. PLoS ONE 6(5): e19653. PDF _just what the title says - AM development, presence threshold 60%, also analyzes richness as a function of threshold_

Ready, J., K. Kaschner, A.B. South, P.D Eastwood, T. Rees, J. Rius, E. Agbayani, S. Kullander and R. Froese (2010). Predicting the distributions of marine organisms at the global scale. Ecological Modelling 221(3): 467-478. PDF _Presents AM; assessing AquaMaps against other presence-only species models_

__Papers based on AquaMaps:__

Jones, M.C., S.R. Dyeb, J.K. Pinnegar and W.W.L. Cheung (2012). Modelling commercial fish distributions: Prediction and assessment using different approaches. Ecological Modelling 225(2012): 133-145. PDF _comparison of species distribution models including AquaMaps, Maxent and the Sea Around Us Project_

Coll, M., C. Piroddi, J. Steenbeek, K. Kaschner, F. Ben Rais Lasram et al. (2010). The biodiversity of the Mediterranean Sea: estimates, patterns, and threats. PLoS ONE 5(8): e11842. PDF _used AquaMaps to predict Med biodiversity.  Also: Threshold = 0._

Martin C.S., Fletcher R., Jones M.C., Kaschner K., Sullivan E., Tittensor D.P., Mcowen C., Geffert J.L., van Bochove J.W., Thomas H., Blyth S., Ravillious C., Tolley M., Stanwell-Smith D. (2014). Manual of marine and coastal datasets of biodiversity importance. May 2014 release. Cambridge (UK): UNEP World Conservation Monitoring Centre. 28 pp. (+ 4 annexes totalling 174 pp. and one e-supplement). PDF _report on marine data sets and data gaps etc, incl both IUCN and AM_

@@

Hurlbert 2007 Species richness, hotspots, and the scale dependence of range maps in ecology and conservation. _mostly rasters of range maps?  "The scale dependence of range-map accuracy poses clear limitations on braod-scale ecological analyses and conservation assessments.  ... we provide guidance about the approriate scale of their use_

Jetz 2008 Ecological Correlates and Conservation Implications of Overestimating Species Geographic Ranges _EOO maps are usually highly interpolated and overestimate small-scale occurrence, which may bias research outcomes_

Pimm 2014 The biodiversity of species and their rates of extinction, distribtuion, and protection. _uses range maps to show biodiversity areas; may use IUCN range maps.  Also discusses gaps and possible things that can be done about them._

Rondinini 2006 Tradeoffs of different types of species occurrence data for use in systematic conservation planning _compares point locality, range maps, and distribution models in terms of omission and commission errors; also outlines Extent of Occurrence and Area of Occupancy distinctions._

@@

@@

* IUCN reference:
    * https://www.conservationtraining.org/course/view.php?id=217&lang=en
    * http://www.iucnredlist.org/technical-documents/red-list-training/iucnspatialresources
    * IUCN Red List - accessed 12/21/2015

* AquaMaps reference, and accessed date?

### 2015-12-29 ms_6_supplemental.Rmd

@@

@@

>>>>

### illustrative maps for different quadrants and different mechanistic problems

@@

___Red List inclusion:___

![](figures/bar_by_taxa_assessed.png)

* _All IUCN-mapped species are also included in Red List species, but only `r p_spp_am_rl`% of AquaMaps species._

![](figures/barchart_spp_risk_quads.png)

* Breaking down the quadrants by IUCN extinction risk categories (FIG 3c), we found that species with higher extinction risk tend to be better aligned between the two datasets, perhaps correlated to increased expert scrutiny.  Does higher perceived risk lead to increased attention, and thus better understanding of species distribution? Or conversely, does increased attention to species distribution reveal more species at risk?  Likely both mechanisms are at play on a case-by-case basis, depending on the species' taxon and region. _does this argument bear up to closer scrutiny? CR isn't dominated by Q1 any more_

@@

## AquaMaps: Effect of changing "presence" threshold on apparent distribution

### AquaMaps presence threshold analysis - move to SOM

For our comparisons of global distribution of represented biodiversity and spatial alignment between datasets, we considered "present" to be any cell with a non-zero probability of occurrence, to best approximate the "extent of occurrence" as generally indicated by IUCN maps.  To examine the effect of different presence threshold selections on the represented range of a species, we varied the threshold from 0.05 to 1.00 and calculated the average species range relative to a zero threshold.

### Figure 5 (a, b):

![](figures/AMThreshold.png)

![](figures/ribbon_AMThreshold.png)

>AquaMaps distribution map extent remaining after applying a presence threshold.  (a) A 40% threshold applied to all species in the AquaMaps dataset shows a mean loss of XXX, with a wide distribution in which some species lose nearly all of their apparent range.  (b) Mean (median) remaining extent at increments of presence threshold. Dark grey ribbon includes 25% to 75% quantiles, while light grey ribbon includes 9% to 91% quantiles.

AquaMaps distribution maps indicate "probability of occurrence" within each 0.5° cell, with values ranging from zero to one, rather than a simple present/absent value as indicated by IUCN maps.  Many studies convert this AquaMaps probability to a simple presence value by assigning a threshold value (REF references here).  A higher threshold constrains an analysis to cells with near certainty of occurrence, while a low threshold captures larger areas of increasingly marginal suitability.

At a presence threshold of 40%, as used in the Ocean Health Index Species subgoal, the bulk of AquaMaps species suffer a significant decrease in represented range, and some species lose nearly their entire range.  Incrementing the presence threshold from 0.00 to 1.00 for the entire AquaMaps dataset, the shallow downward trend indicates a low but consistent sensitivity to threshold choice, with no surprising tradeoffs that could suggest an "optimal" threshold.  This pattern may not hold true for all subsets of AquaMaps species, however, whether subsetting by taxa or by georegion.

-----
